\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-15 latex/article_vol2.tex

\subsection{Getting the full cleanR workhorse on your side}

\subsection{A worked example}

@@

\subsubsection{Dropping the figures}
Though figures give a nice overview of each variable, they are also quite heavy objects in terms of memory allocation. Therefore, it might be beneficial to not include figures in the \R{cleanR} outputs for very large datasets. This is controlled via the \R{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check")
\end{Sinput}
\end{Schunk}

@@

>>>>

\subsection{Getting the full cleanR machinery on your side}

\subsection{A worked example}
\begin{figure}
\begin{center}
\includegraphics[scale=0.2]{mosaicPlotExample.pdf}
\end{center}
\label{fig:mosaicPlot}
\caption{\hl{something, note that code for producing this plot is available in ./latex/codeForArticle.R}}
\end{figure}

Now, let's say we wish to perform quality control on a slightly more complicated dataset than \R{toyData}. \hl{something about which dataset}. We will build four new functions and show both how they can be used interactively and how they can be integrated with the \R{clean()} function. These four new functions are:
\begin{description}
\item[\R{isID}] A new \R{checkFunction} intended for use in the precheck-step. This function checks whether a variable consists exclusively long ($> 10$) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.  
\item[\R{mosaicVisual}] A new \R{visualFunction} that produces so-called mosaic plots. This function will be used in the \textit{visualize} step of \R{clean()}.
\item[\R{countZeros}] A new \R{summaryFunction} that counts the number of occurrences of the value \R{0} in a variable. This function will be used in the \textit{summarize} step of \R{clean()}.
\item[\R{identifyColon}] A new \R{checkFunction} that flags variables that colons that appear before and after alphanumerical characters, e.g. practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \R{clean()}. 
\end{description}
First, let's define the \R{isID} function. We do not particularly wish to use this function interactively, so we will stick to the minimal requirements of a \R{checkFunction} used in \R{check()}. The function might then be defined by
\begin{Schunk}
\begin{Sinput}
isID <- function(v, nMax = NULL, ...) {
	out <- list(problem = FALSE, message = "")
	if (class(v) %in% c("character", "factor", "labelled", "numeric", "integer")) {
		v <- as.character(v)
		lengths <- c(nchar(v))
		if (all(lengths > 10) & length(unique(lengths)) == 1) {
			out$problem <- TRUE
			out$message <- "Warning: This variable seems to contain ID codes."
		}
	}
	out
}
\end{Sinput}
\end{Schunk} 
This is essentially all we need to do in order to include this function as precheck-function in \R{clean()}, so we will leave this function as is and move on to the next function, namely \R{mosaicVisual}. We will define this function such that it gets the full \R{cleanR} functionality. This can be done using the code 
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- function(v, vnam, doEval) {
   thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
   if (doEval) {
    return(eval(thisCall))
   } else return(deparse(thisCall))
 }
\end{Sinput}
\end{Schunk}
This function can now be called directly or used in \R{clean()}. We will return to its usage in \R{clean()} below. Depending on the argument \R{doEval}, either a text string with code or a plot is produced. The plot resulting from the following call is found in Figure~\ref{fig:mosaicPlot}:
\begin{Schunk}
\begin{Sinput}
mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)   $
\end{Sinput}
\end{Schunk}
\hl{remove \$.}
Even though \R{mosaicVisual}, as written above, follows the style of a \R{visualFunction}, it is not yet truly one and therefore, it will not appear in a \R{allVisualFunctions()} call. In order to get this functionality, we need to change its object class. This can be done by writing
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- visualFunction(mosaicVisual, 
 				description = "Mosaic plots using graphics",
                 classes = allClasses())
\end{Sinput}
\end{Schunk}
Here, we use the function \R{allClasses()} to quickly obtain a vector of all the seven variable classes addressed in \R{cleanR}. Note that if R{mosaicVisual} has been an S3 generic function, this argument could have been left as \R{NULL} and then the classes for which methods are available would be added automatically.

As \R{mosaicVisual} is now a full blood \R{visualFunction}, it will also be included in the \R{allVisualFunctions()} output table:
\begin{Schunk}
\begin{Sinput}
> allVisualFunctions()
\end{Sinput}
\begin{Soutput}
------------------------------------------------------------------------
name           description                   classes                    
-------------- ----------------------------- ---------------------------
mosaicVisual   Mosaic plots using graphics   character, Date, factor,   
                                             integer, labelled, logical,
                                             numeric

basicVisual    Histograms and barplots using                            
               graphics

standardVisual Histograms and barplots using                            
               ggplot2                                                  
------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}
\hl{Redo this output table when makeXFunction issues are fixed.}

Now, we are done with the definition of \R{mosaicVisual} and we can turn to the next function in line, \R{countZeros}. This \R{summaryFunction} in spe is defined in the following lines of code:
\begin{Schunk}
\begin{Sinput}
countZeros <- function(v, ...) {
	res <- length(which(v == 0))
	summaryResult(list(feature = "No. zeros", result = res, value = res))
}
\end{Sinput}
\end{Schunk}
Note that as this function computes an integer (the number of zeros), there is no difference between the entires \R{\$result} and \R{\$value}. If, on the other hand, the result had been a character string, extra formatting might be required in the \R{\$result} entry (such as escaping of quotation marks), and in this scenario, the two entries would then differ. As the result is returned as a \R{summaryResult} object, a printing method is automatically called when \R{countZeros} is used interactively:
\begin{Schunk}
\begin{Sinput}
> countZeros(c(rep(0, 5), 1:100))
\end{Sinput}
\begin{Soutput}
No. zeros: 5
\end{Soutput}
\end{Schunk}
As in the above, we change the class of this function in order to make it appear in \R{allSummaryFunctions()} calls. But now we wish to emphasize that the function is not intended to be called on all variable types, as zeros have different roles in \R{Date}s and in \R{logical} variables:
\begin{Schunk}
\begin{Sinput}
> countZeros <- summaryFunction(countZeros,
	description = "Count number of zeros",
	classes = c("character", "factor", "integer", 
	  "labelled", "numeric"))
\end{Sinput}
\end{Schunk}

The last function mentioned above is \R{identifyColon()}. We define it using the helper function \R{messageGenerator} to obtain a properly escaped message, and we use \R{checkResult} to make it appear in \R{allCheckFunction()} calls: 
\begin{Schunk}
\begin{Sinput}
identifyColon <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

  if (length(problemValues) > 0) {
    problem <- TRUE 
  }

  problemStatus <- list(problem = problem, 
                        problemValues = problemValues)
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)

  checkResult(list(problem = problem, 
                   message = outMessage,
                   problemValues = problemValues))
}

identifyColon <- checkFunction(identifyColon, 
    description = "Identify non-suffixed nor -prefixed colons",
    classes = c("character", "factor", "labelled"))

\end{Sinput}
\end{Schunk}
As with the previous two functions, we also change its class. Note, however, that for \R{checkFunction}s, the function description will appear in the document produced by \R{clean()} (in the \textit{Data cleaning summary} section), so now this is not only done for the sake of the \R{allCheckFunctions()} output.\\

Now, we are ready to use these new functions in a \R{clean()} call. \hl{do this... what dataset should we use here? Include output in appendix, maybe.}. The extended \R{cleanR} output document should have the following modifications, relative to the standard \R{cleanR} output:
\begin{itemize}
\item We want to add the new pre-check function, \R{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \R{countZeros}, to be added to the summaries performed on all variable types but \R{Date} and \R{logical}.
\item We want the new check function, \R{identifyColon}, to be added to the checks performed on \R{character}, \R{factor} and \R{labelled} variables.
\end{itemize}
These options are specified as follows:
\begin{Schunk}
\begin{Sinput}
> clean(dataSet, 
      preChecks = c("isKey", "isEmpty", "isID"),
      allVisuals = "mosaicVisual",
      characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
      factorSummaries = c(defaultFactorSummaries(), "countZeros"),
      labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
      numericSummaries = c(defaultNumericSummaries(), "countZeros"),
      integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
      characterChecks = c(defaultCharacterChecks(), "identifyColon"),
      factorChecks = c(defaultFactorChecks(), "identifyColon"),
      labelledCheck = c(defaultLabelledChecks(), "identifyColon"))
\end{Sinput}
\end{Schunk}
The outputted document is found in Appendix \hl{NUMTWO}. \hl{Comment. Make appendix. Remember to change dataSet to something else, when we decide what data to use here.}

@@

\subsubsection{Dropping the figures}
Though figures give a nice overview of each variable, they are also quite heavy objects in terms of memory allocation. Therefore, it might be beneficial to not include figures in the \R{cleanR} outputs for very large datasets. This is controlled via the \R{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}
\hl{blabla, end?}

@@

\section{Appendix NUMTWO}
\hl{Data cleaning with extension here}
